---
title: "Football 2021 Model"
author: "Donovan Robertson"
date: "12/04/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Files

Load the packages to use in R Studio


```{r}
library(dplyr)
library(skellam)
library(ggplot2)
library(purrr)
library(tidyr)
library(MASS)
library(pscl)
library(lmtest)
library(nonnest2)
```


## Select the EPL data


```{r}
options(stringsAsFactors = FALSE)
footballdata<-read.csv("FootballData.csv")

epldata<-subset(footballdata, Div=="E0", select=c(4:7))
colnames(epldata)<-c("home", "away", "homegoals", "awaygoals")
epldata$homegoals=as.numeric(as.character(epldata$homegoals))
epldata$awaygoals=as.numeric(as.character(epldata$awaygoals))
```


# Comparison of EPL average goals scored vs against

```{r}
data.frame(average_home_goals=mean(epldata$homegoals),
           average_away_goals=mean(epldata$awaygoals))
```


#Distribution of goals

```{r}
mean(epldata$homegoals)
```

```{r}
epldata %>% count(epldata$homegoals)
```


```{r}
epldata %>% count(epldata$awaygoals)
```





#Histograms
```{r}
hist(epldata$homegoals, breaks=10,  main = "Histogram of EPL Home Goals", xlab="Number of Home Team Goals", freq=FALSE,xaxp=c(0,10,10))
lines(0:max(epldata$homegoals), dpois(0:max(epldata$homegoals), mean(epldata$homegoals)), col='red')
```

# Comparison of EPL average goals scored vs against

```{r}
data.frame(average_home_goals=mean(epldata$homegoals),
           average_away_goals=mean(epldata$awaygoals))
```

```{r}
data.frame(variance_home_goals=var(epldata$homegoals),
           variance_away_goals=var(epldata$awaygoals))
```




```{r}
x=(0:9)
dpois(x,1.334416,FALSE)
```




```{r}
hist(epldata$awaygoals, breaks=8, main= "Histogram of EPL Away Goals", xlab="Number of Away Team Goals", freq=FALSE, xaxp=c(0,8,8))
lines(0:max(epldata$awaygoals), dpois(0:max(epldata$awaygoals), mean(epldata$awaygoals)), col='red')
```

# Comparison of EPL average goals scored vs against

```{r}
data.frame(average_home_goals=mean(epldata$homegoals),
           average_away_goals=mean(epldata$awaygoals))
```


```{r}
x2=(0:9)
dpois(x2,1.340909,FALSE)
```


```{r}
var(epldata$homegoals)
var(epldata$awaygoals)
```

```{r}
AIC(eplpoisson_model, eplzeroinfl_model)
```



#EPL Poisson GLM Model 
```{r}
eplpoisson_model<-
  rbind(
    data.frame(goals=epldata$homegoals,
               team=epldata$home,
               opponent=epldata$away,
               home=1),
    data.frame(goals=epldata$awaygoals,
               team=epldata$away,
               opponent=epldata$home,
               home=0)) %>%
  glm(goals ~ home + team + opponent, family=poisson(link=log), data=.)

summary(eplpoisson_model)
```

#Model Checking

```{r}
fitted.values<-eplpoisson_model$fitted.values
standard.residuals<-(eplpoisson_model$residuals)
plot(fitted.values,standard.residuals, ylab='Standardised Residuals', xlab='Fitted Values', abline(0,0,col='blue'), ylim=c(-2,8), main='Residual Plot for EPL Poisson GLM')
hist(standard.residuals, xlim=c(-1.75,7), main='Histogram of Standard Residuals', xlab='Standard Residuals')
qqnorm(standard.residuals, ylim=c(-1,7))
abline(0,1,col='blue')
```

```{r}
dput(coef(eplzeroinfl_model,"count"))
dput(coef(eplzeroinfl_model, "zero"))
```







```{r}
qqnorm(epldata$awaygoals)
qqline(epldata$awaygoals, col='red')
```


#Negative Biomial Model


glm.nb(goals ~ home + team + opponent, family=negative.binomial(link=log), data=.)


```{r}
eplnegativebinomial_model<-
  rbind(
    data.frame(goals=epldata$homegoals,
               team=epldata$home,
               opponent=epldata$away,
               home=1),
    data.frame(goals=epldata$awaygoals,
               team=epldata$away,
               opponent=epldata$home,
               home=0)) %>%
  glm.nb(goals ~ home + team + opponent, data=.)

summary(eplnegativebinomial_model)
```

```{r}
fitted.values2<-eplnegativebinomial_model$fitted.values
standard.residuals2<-(eplnegativebinomial_model$residuals)
plot(fitted.values2,standard.residuals2)
hist(standard.residuals2)
qqnorm(standard.residuals2)
abline(0,1)
```



##Poisson Zero Inflated Model 

```{r}
eplzeroinfl_model<-
  rbind(
    data.frame(goals=epldata$homegoals,
               team=epldata$home,
               opponent=epldata$away,
               home=1),
    data.frame(goals=epldata$awaygoals,
               team=epldata$away,
               opponent=epldata$home,
               home=0)) %>%
 zeroinfl(goals ~ home + team + opponent, family=poisson(link=log), data=.)

summary(eplzeroinfl_model)
```

```{r}
fitted.values3<-eplzeroinfl_model$fitted.values
standard.residuals3<-(eplzeroinfl_model$residuals)
plot(fitted.values3,standard.residuals3, main='Residual Plot for EPL Zero Inflated Poisson GLM', ylab='Standard Residuals', xlab='Fitted Values', abline(0,0, col='blue'), ylim=c(-4,6))
hist(standard.residuals3, main='Histogram of Standard Residuals', xlab='Standard Residuals')
qqnorm(standard.residuals3)
abline(0,1,col='blue')
```


```{r}
eplquasi_model<-
  rbind(
    data.frame(goals=epldata$homegoals,
               team=epldata$home,
               opponent=epldata$away,
               home=1),
    data.frame(goals=epldata$awaygoals,
               team=epldata$away,
               opponent=epldata$home,
               home=0)) %>%
 glm(goals ~ home + team + opponent, family=quasipoisson(link=log), data=.)

summary(eplquasi_model)
```

```{r}
fitted.values4<-eplquasi_model$fitted.values
standard.residuals4<-(eplquasi_model$residuals)
plot(fitted.values4,standard.residuals4, main='Residual Plot for EPL Zero Inflated Poisson GLM', ylab='Standard Residuals', xlab='Fitted Values', abline(0,0, col='blue'), ylim=c(-4,6))
hist(standard.residuals4, main='Histogram of Standard Residuals', xlab='Standard Residuals')
qqnorm(standard.residuals4)
abline(0,1,col='blue')
```

```{r}
lrtest(eplpoisson_model, eplzeroinfl_model)
```

```{r}
vuongtest(eplpoisson_model, eplzeroinfl_model)
```




```{r}
exp(0.1293802366)
exp(-0.2688992893)
1-0.7642202
```

```{r}
0.24947985/0.23358043
```


```{r}
res=resid(eplpoisson_model, type="pear")
sqrt(sum(res^2)/576)
```


```{r}
predict(eplpoisson_model, data.frame(home=1, team="Leeds", opponent="Liverpool"), type="response")
```

```{r}
exp(0.1293802-0.0001433+0.1464553+0.0967824)
0.1293802-0.0001433+0.1464553+0.0967824
```


```{r}
predict(eplpoisson_model, data.frame(home=0, team="Liverpool", opponent="Leeds"), type="response")
```

```{r}
exp(0.1293802+0.2379885+0.3902404)
0.1293802+0.2379885+0.3902404
```


```{r}
predict(eplpoisson_model, data.frame(home=1, team="Arsenal", opponent="Fulham"), type="response")
```

```{r}
exp(0.1293802-0.0001433+0.1609880)
```






```{r}
simulate_match<-function(foot_model, hometeam, awayteam, max_goals=10){
  home_goals_average <-predict(foot_model, 
                               data.frame(home=1, team=hometeam,
                                          opponent=awayteam), type="response")
  away_goals_average<-predict(foot_model,
                              data.frame(home=0, team=awayteam,
                                         opponent=hometeam), type="response")
  dpois(0:max_goals,home_goals_average) %o% dpois(0:max_goals, away_goals_average)}
```

```{r}
simulate_match(eplzeroinfl_model, "Leeds", "Liverpool", max_goals=10) 
```


#EPL Poisson Predictions

```{r}
Everton_Tottenham<-simulate_match(eplpoisson_model, "Everton", "Tottenham", max_goals=10)
sum(Everton_Tottenham[lower.tri(Everton_Tottenham)])
sum(diag(Everton_Tottenham))
sum(Everton_Tottenham[upper.tri(Everton_Tottenham)])
```


```{r}
Newcastle_WestHam<-simulate_match(eplpoisson_model, "Newcastle", "West Ham", max_goals=10)
sum(Newcastle_WestHam[lower.tri(Newcastle_WestHam)])
sum(diag(Newcastle_WestHam))
sum(Newcastle_WestHam[upper.tri(Newcastle_WestHam)])
```


```{r}
Wolves_SheffieldUnited<-simulate_match(eplpoisson_model, "Wolves", "Sheffield United", max_goals=10)
sum(Wolves_SheffieldUnited[lower.tri(Wolves_SheffieldUnited)])
sum(diag(Wolves_SheffieldUnited))
sum(Wolves_SheffieldUnited[upper.tri(Wolves_SheffieldUnited)])
```


```{r}
Arsenal_Fulham<-simulate_match(eplpoisson_model, "Arsenal", "Fulham", max_goals=10)
sum(Arsenal_Fulham[lower.tri(Arsenal_Fulham)])
sum(diag(Arsenal_Fulham))
sum(Arsenal_Fulham[upper.tri(Arsenal_Fulham)])
```


```{r}
ManUnited_Burnley<-simulate_match(eplpoisson_model, "Man United", "Burnley", max_goals=10)
sum(ManUnited_Burnley[lower.tri(ManUnited_Burnley)])
sum(diag(ManUnited_Burnley))
sum(ManUnited_Burnley[upper.tri(ManUnited_Burnley)])
```




```{r}
Leeds_Liverpool<-simulate_match(eplpoisson_model, "Leeds", "Liverpool", max_goals=10)
sum(Leeds_Liverpool[lower.tri(Leeds_Liverpool)])
sum(diag(Leeds_Liverpool))
sum(Leeds_Liverpool[upper.tri(Leeds_Liverpool)])
```

```{r}
Chelsea_Brighton<-simulate_match(eplpoisson_model, "Chelsea", "Brighton", max_goals=10)
sum(Chelsea_Brighton[lower.tri(Chelsea_Brighton)])
sum(diag(Chelsea_Brighton))
sum(Chelsea_Brighton[upper.tri(Chelsea_Brighton)])
```



#EPL Zero Inflation Predictions


```{r}
Everton_Tottenham<-simulate_match(eplzeroinfl_model, "Everton", "Tottenham", max_goals=10)
sum(Everton_Tottenham[lower.tri(Everton_Tottenham)])
sum(diag(Everton_Tottenham))
sum(Everton_Tottenham[upper.tri(Everton_Tottenham)])
```


```{r}
Newcastle_WestHam<-simulate_match(eplzeroinfl_model, "Newcastle", "West Ham", max_goals=10)
sum(Newcastle_WestHam[lower.tri(Newcastle_WestHam)])
sum(diag(Newcastle_WestHam))
sum(Newcastle_WestHam[upper.tri(Newcastle_WestHam)])
```


```{r}
Wolves_SheffieldUnited<-simulate_match(eplzeroinfl_model, "Wolves", "Sheffield United", max_goals=10)
sum(Wolves_SheffieldUnited[lower.tri(Wolves_SheffieldUnited)])
sum(diag(Wolves_SheffieldUnited))
sum(Wolves_SheffieldUnited[upper.tri(Wolves_SheffieldUnited)])
```


```{r}
Arsenal_Fulham<-simulate_match(eplzeroinfl_model, "Arsenal", "Fulham", max_goals=10)
sum(Arsenal_Fulham[lower.tri(Arsenal_Fulham)])
sum(diag(Arsenal_Fulham))
sum(Arsenal_Fulham[upper.tri(Arsenal_Fulham)])
```


```{r}
ManUnited_Burnley<-simulate_match(eplzeroinfl_model, "Man United", "Burnley", max_goals=10)
sum(ManUnited_Burnley[lower.tri(ManUnited_Burnley)])
sum(diag(ManUnited_Burnley))
sum(ManUnited_Burnley[upper.tri(ManUnited_Burnley)])
```


```{r}
Leeds_Liverpool<-simulate_match(eplzeroinfl_model, "Leeds", "Liverpool", max_goals=10)
sum(Leeds_Liverpool[lower.tri(Leeds_Liverpool)])
sum(diag(Leeds_Liverpool))
sum(Leeds_Liverpool[upper.tri(Leeds_Liverpool)])
```


```{r}
Chelsea_Brighton<-simulate_match(eplzeroinfl_model, "Chelsea", "Brighton", max_goals=10)
sum(Chelsea_Brighton[lower.tri(Chelsea_Brighton)])
sum(diag(Chelsea_Brighton))
sum(Chelsea_Brighton[upper.tri(Chelsea_Brighton)])
```



